\section*{Data and Code Availability}

All datasets used in this study are public molecular property datasets. The experimental workflow, preprocessing scripts, split-generation scripts, model-training scripts, audit scripts, result tables, manuscript figures, and LaTeX source files are maintained in the public reproducibility repository: \url{https://github.com/DCarchimonde/split-diagnostic-molecular-benchmark}. The analysis can be reproduced by following the commands listed in Appendix~\ref{app:reproducibility}.

\section*{Competing Interests}

The authors declare no competing interests.

\section*{Author Contributions}

S.T. conceptualized the study, designed the benchmark audit, implemented the experimental workflow, analyzed the results, drafted the manuscript, and serves as the corresponding author. Y.W. reviewed the methodology, checked the experimental design and manuscript consistency, provided technical feedback, and contributed to manuscript revision. Both authors approved the final manuscript.

\section*{Declaration of AI-Assisted Technologies in the Manuscript Preparation Process}

During the preparation of this work, the authors used an AI-assisted tool to support language editing, manuscript organization, LaTeX formatting, and reproducibility-material preparation. The authors reviewed and edited all content, verified the references, code, figures, and numerical results, and take full responsibility for the submitted manuscript.

\section*{Funding}

The authors received no specific funding for this work.

\section*{Acknowledgements}

The authors thank the maintainers of the public molecular datasets and open-source cheminformatics and machine-learning libraries used in this work.
