\section{Discussion}

This study shows that molecular property prediction benchmarks should be interpreted as a joint product of the model, the dataset, and the train/test split. Random splitting, ordinary scaffold splitting, and target-balanced scaffold splitting do not simply provide interchangeable test sets. They produce different chemical similarity patterns, target distributions, scaffold compositions, performance gaps, and sometimes model rankings. Therefore, a benchmark score without split diagnostics is only partially interpretable.

The most direct implication is that random-split performance should be treated cautiously in molecular property prediction. In this audit, random splits generally produced higher maximum test-to-train Tanimoto similarity than scaffold-based splits. This does not mean random splits are useless; they remain useful for measuring interpolation-like performance under chemically similar train/test distributions. However, they should not be interpreted as evidence of scaffold-level chemical generalization unless accompanied by similarity diagnostics.

The second implication is that ordinary scaffold splitting is useful but not self-explanatory. It successfully removes train/test scaffold overlap, which is valuable. At the same time, it can induce target shift and scaffold concentration. ESOL and FreeSolv illustrate this problem clearly: the ordinary scaffold split holds out very few scaffold groups and creates large target gaps. In such cases, a performance drop under scaffold splitting may reflect a mixture of scaffold generalization, target distribution shift, and small-test-set instability. Reporting scaffold statistics and target gaps makes this ambiguity visible rather than hidden.

The target-balanced scaffold split helps clarify this ambiguity. When gaps shrink after target balancing, as seen for several ESOL tree-based results and some classification models, part of the ordinary scaffold gap is plausibly associated with target shift. When gaps remain or increase, as observed in HIV, the split may be probing a different difficulty regime. This is why target-balanced scaffold splitting should be interpreted as a diagnostic counterfactual rather than a universal benchmark replacement. Its value lies in making the benchmark question more explicit.

The broader message is practical. AIDD benchmark papers should not only ask which model obtains the highest score. They should also ask what kind of test set produced that score. At minimum, benchmark reports should include the split protocol, target distribution diagnostics, scaffold statistics, test-to-train chemical similarity, and model-ranking stability. These additions do not require complex new models, but they can substantially improve the transparency of model comparisons.

\section{Limitations}

This study has several limitations. First, the experiments use fingerprint-based classical baselines rather than graph neural networks or transformer-based molecular encoders. This was intentional: lightweight baselines isolate split effects and improve reproducibility. Nevertheless, future work should test whether the same diagnostic patterns hold for deep molecular representations. Second, the target-balanced scaffold split is heuristic. It is used here as a diagnostic counterfactual, not as a new universal split standard. Third, the benchmark includes six public datasets, which provide a focused audit across classification and regression tasks but do not cover all AIDD scenarios. Fourth, no prospective or time-based validation is included. Finally, small datasets such as FreeSolv may be structurally constrained by scaffold concentration, making it difficult to construct a split that is simultaneously scaffold-separated, target-balanced, and statistically stable.

\section{Conclusion}

Molecular property prediction benchmarks should treat data splitting as an object of diagnosis. Random splits can produce chemically closer test sets and optimistic performance estimates. Ordinary scaffold splits remove scaffold overlap but can introduce target shift and scaffold concentration. Target-balanced scaffold splits help identify whether scaffold gaps persist after reducing target shift, but their effects remain dataset- and model-dependent. A split-diagnostic audit that reports performance together with target distribution shift, scaffold statistics, chemical similarity, outlier cases, and model-ranking stability provides a more transparent basis for evaluating generalization in AI-driven drug discovery.
